---
title: "Practicum I CS5200"
author field: Pratik Chatterjee
date: ' Summer Full 2023'
output:
  html_document:
    df_print: paged
---

# Pratik Chatterjee
## chatterjee.pr@northeastern.edu


# Settings freemysqlhosting.net (max 5MB)
# Specify the database connection details
## Connect to database
```{r DatabaseCreation}
library(RMySQL)
library(DBI)
#Settings freemysqlhosting.net (max 5MB)
db_name_fh <- "sql7628203"
db_user_fh <- "sql7628203"
db_host_fh <- "sql7.freemysqlhosting.net"
db_pwd_fh <- "NPqraIa1dg"
db_port_fh <- 3306

# 3. Connect to remote server database
mydb.fh <- dbConnect(RMySQL::MySQL(), user = db_user_fh, password = db_pwd_fh, dbname = db_name_fh, host = db_host_fh, port = db_port_fh)

mydb <- mydb.fh
```

## Create Database
## Drop existing tables if they exist
```{sql dropstrikes,connection=mydb}
DROP TABLE IF EXISTS strikes
```

```{sql dropflights,connection=mydb}
DROP TABLE IF EXISTS flights
```

```{sql dropairports,connection=mydb}
DROP TABLE IF EXISTS airports
```

```{sql dropconditions,connection=mydb}
DROP TABLE IF EXISTS conditions
```

# Create the airports table
```{sql airporttable, connection=mydb}
# Create the airports table
CREATE TABLE airports (
                 aid INT PRIMARY KEY,
                 airportState TEXT,
                 airportCode TEXT
               );
```

# Create the conditions table
```{sql conditionstable, connection=mydb}
# Create the conditions table
CREATE TABLE conditions (
                 cid INT PRIMARY KEY,
                 sky_condition TEXT,
                 explanation TEXT
               );
```

# Create the flights table
```{sql flightstable, connection=mydb}
# Create the flights table
CREATE TABLE flights (
                 fid INT PRIMARY KEY,
                 date DATE,
                 origin INT,
                 airline TEXT,
                 aircraft TEXT,
                 altitude INT,
                 heavy BOOLEAN,
                 FOREIGN KEY (origin) REFERENCES airports (aid)
               );

```

# Create the strikes table
```{sql strikestable, connection=mydb}
# Create the strikes table
CREATE TABLE strikes (
                 sid INT PRIMARY KEY,
                 fid INT,
                 numbirds INT,
                 impact TEXT,
                 damage BOOLEAN,
                 altitude INT,
                 conditions INT,
                 FOREIGN KEY (fid) REFERENCES flights (fid),
                 FOREIGN KEY (conditions) REFERENCES conditions (cid)
               );
```

# csv file
# Specify the path and name of the CSV file containing the bird strike data
```{r importCSVfile}
# csv file
csv_file <- "BirdStrikesData-V2.csv"
```

# read csv file
# Read the CSV file into a data frame
```{r CSVfiletodataframe}
# read csv file
bds.raw <- read.csv(file = csv_file, header = T, stringsAsFactors = F)
```

```{r dateformatconversion}
# Convert the flight_date column to date format
bds.raw$flight_date <- as.Date(bds.raw$flight_date, format = "%m/%d/%Y")
```


```{r first10data}
names(bds.raw) <- c('rid','aircraft','airport',	'model','wildlife_struck',	'impact',	'flight_date',	'damage',	'airline', 'origin','flight_phase',	'remains_collected_flag',	'Remarks','wildlife_size',	'sky_conditions', 'species','pilot_warned_flag', 'altitude_ft',	'heavy_flag')

head(bds.raw, 10)

```

#Create airports df for MySQL db
```{r airportsdataframe}
aid <- NA
airportCode <- NA
airportState <- ifelse(is.na(bds.raw$origin), "Unknown", bds.raw$origin)
airports_df <- data.frame(aid, airportCode, airportState)
```

#Assign synthetic key values for airports table
```{r synthetickeyvalueairport}
aid <- row.names(airports_df)
airports_df <- data.frame(aid, airportCode, airportState)
```

# Prepare the airports data frame for insertion into the MySQL database
```{r airportinsertion}
dbWriteTable(mydb, "airports", airports_df, row.names = FALSE, append = TRUE)

```
```{sql connection=mydb}
SELECT * FROM airports LIMIT 10;
```
#Create conditions df for MySQL db
```{r conditionsdataframe}
cid <- NA
sky_condition <- bds.raw$sky_conditions 
explanation <- NA
conditions_df <- data.frame(cid, sky_condition, explanation)
```

#Assign synthetic key values for conditions table
```{r synthetickeyvalueconditions}
cid <- row.names(conditions_df)
conditions_df <- data.frame(cid, sky_condition, explanation)
```

# Prepare the conditions data frame for insertion into the MySQL database
```{r conditionsinsertion}
dbWriteTable(mydb, "conditions", conditions_df, row.names = FALSE, append = TRUE)
```
```{sql connection=mydb}
SELECT * FROM conditions LIMIT 10;
```

#Create flights df for MySQL db
```{r flightsdataframe}
fid <- bds.raw$rid
date <- as.Date(bds.raw$flight_date, format = "%m/%d/%Y") 
origin <- airports_df$aid
airline <- bds.raw$airline
aircraft <- bds.raw$aircraft
altitude <- bds.raw$altitude_ft
heavy <- bds.raw$heavy_flag
flights_df <- data.frame(fid, date, origin, airline, aircraft, altitude, heavy)
```

# Prepare the flights data frame for insertion into the MySQL database
```{r flightsinsertion}
dbWriteTable(mydb, "flights", flights_df, row.names = FALSE, append = TRUE)
```
```{sql connection=mydb}
SELECT * FROM flights LIMIT 10;
```

#Create strikes df for MySQL db
```{r strikessdataframe}
sid <- NA
fid <- bds.raw$rid
altitude <- bds.raw$altitude_ft
conditions <- conditions_df$cid
numbirds <- bds.raw$wildlife_struck
damage <- bds.raw$damage
impact <- bds.raw$impact
strikes_df <- data.frame(sid, fid, altitude, conditions, numbirds, damage, impact)
```

#Assign synthetic key values for strikes table
```{r synthetickeyvaluestrikes}
sid <- row.names(strikes_df)
strikes_df <- data.frame(sid, fid, altitude, conditions, numbirds, damage, impact)
```

# Prepare the strikes data frame for insertion into the MySQL database
```{r strikesinsertion}
dbWriteTable(mydb, "strikes", strikes_df, row.names = FALSE, append = TRUE)
```

```{sql connection=mydb}
SELECT * FROM strikes LIMIT 10;
```

## 8.Query to find the top 10 states with the greatest number of bird strike incidents
```{sql Task8SQLQuery, connection=mydb}
SELECT airportState AS state, COUNT(*) AS incidents
FROM strikes
JOIN flights ON strikes.fid = flights.fid
JOIN airports ON flights.origin = airports.aid
GROUP BY airportState
ORDER BY incidents DESC
LIMIT 10;
```

## 9.Query to find the airlines with an above average number of bird strike incidents
```{sql Task9SQLQuery, connection=mydb}
SELECT airline, COUNT(*) AS incidents
FROM strikes
JOIN flights ON strikes.fid = flights.fid
GROUP BY airline
HAVING COUNT(*) > (SELECT AVG(incident_count) FROM (SELECT COUNT(*) AS incident_count FROM strikes JOIN flights ON strikes.fid = flights.fid GROUP BY airline) AS subquery)
ORDER BY incidents DESC;
```


## 10.Query to find the (total) number of birds that struck aircraft by month
```{r Task10SQLQuery}
query<-"SELECT EXTRACT(MONTH FROM flights.date) AS month, SUM(numbirds) AS total_bird_strikes
FROM strikes
JOIN flights ON strikes.fid = flights.fid
WHERE 
numbirds IS NOT NULL AND flights.date IS NOT NULL
GROUP BY month
ORDER BY month"
        
result<- dbGetQuery(mydb,query)
head(result, 6)
```
## 11.Plot a column chart to visualize the number of birds struck by aircraft monthly
```{r Task11SQLQuery}
# Set the plot area size
par(mar = c(5, 4, 4, 4))

# Define the desired y-axis limit
ylim <- c(0, 14000)

# Create the column chart with better spacing and font size
barplot(result$total_bird_strikes, names.arg = result$month, horiz = FALSE,
        col = "blue", main = "No. of Birds Struck by Aircraft Monthwise",
        xlab = "Months", ylab = "No. of Birds", cex.names = 0.7, ylim = ylim)

# Add data labels with adjusted position and font size
text(x = seq_along(result$total_bird_strikes), y = result$total_bird_strikes,
     labels = result$total_bird_strikes, pos = 3, cex = 0.8)

# Adjust the plot margins
par(mar = c(5, 4, 4, 4) + 1)
```
```{sql dropprocedure, connection=mydb}
DROP PROCEDURE IF EXISTS InsertStrike;

```

## 12.Create a stored procedure to add a strike record to the database
```{sql Task12StoredProcedurecreation, connection=mydb}
CREATE PROCEDURE InsertStrike(
    IN new_airportState TEXT,
    IN new_airportCode TEXT,
    IN new_airline TEXT,
    IN new_aircraft TEXT,
    IN new_flightDate DATE,
    IN new_skyCondition TEXT,
    IN new_explanation TEXT,
    IN new_numbirds INT,
    IN new_impact TEXT,
    IN new_damage BOOLEAN,
    IN new_altitude INT
)
BEGIN
    DECLARE new_aid INT;
    DECLARE new_fid INT;
    DECLARE new_cid INT;
    
    -- Check if the airport already exists
    SELECT aid INTO new_aid FROM airports WHERE airportCode = new_airportCode;
    
    IF new_aid IS NULL THEN
        -- Insert new airport if it doesn't exist
        INSERT INTO airports (airportState, airportCode) VALUES (new_airportState, new_airportCode);
        SET new_aid = LAST_INSERT_ID();
    END IF;
    
    -- Insert new flight
    INSERT INTO flights (date, origin, airline, aircraft, altitude) 
    VALUES (new_flightDate, new_aid, new_airline, new_aircraft, new_altitude);
    SET new_fid = LAST_INSERT_ID();
    
    -- Check if the conditions already exist
    SELECT cid INTO new_cid FROM conditions WHERE sky_condition = new_skyCondition AND explanation = new_explanation;
    
    IF new_cid IS NULL THEN
        -- Insert new conditions if they don't exist
        INSERT INTO conditions (sky_condition, explanation) VALUES (new_skyCondition, new_explanation);
        SET new_cid = LAST_INSERT_ID();
    END IF;
    
    -- Insert new strike
    INSERT INTO strikes (fid, numbirds, impact, damage, altitude, conditions) 
    VALUES (new_fid, new_numbirds, new_impact, new_damage, new_altitude, new_cid);
    
    SELECT * FROM strikes WHERE sid = LAST_INSERT_ID();
END 
```


#Call the stored procedure and insert a new strike
##CALL InsertStrike(1, 5, 'Engine Damage', TRUE, 10000, 1);
```{sql storedprocinsertion, connection=mydb}
#Call the stored procedure and insert a new strike
CALL InsertStrike('New York', 'JFK', 'Delta Airlines', 'Boeing 737', '2023-06-22', 'Clear', 'Clear sky conditions', 3, 'Engine Damage', TRUE, 10000);
```


```{r Disconnectingdatabase}
dbDisconnect(mydb)
```



